Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-19 paper-amia-summit-2018/amia_summit_2018.tex
update introduction

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. For classifying temporally ordered sequences, sequence classification methods are utilized. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}. In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease. Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.

@@

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information was learned from the dataset, and use that as a "model" to classify sequences.

@@

In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

%This work is the next step in a series of experiments to develop the [informatics based methods described above].

In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this work, we propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.

@@

Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.  A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively. Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors. For this reason, we only consider those models that can capture the past events for the classification of behavior sequences.\\

@@

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Mathematically, HMM can be defined as the following: $\lambda = (N, M, A, B, \pi)$
\begin{itemize}
\item N is a number of hidden states in the model
\item M is a number of distinct observations symbols per state, i.e. the discrete vocabulary size
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
HMM is also represented by a compact notation: $\lambda = (A, B, \pi)$ after excluding the structure parameters M and N. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence. Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful). This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\subsection*{\textit{Most likely communication sequences}}
Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. In successful sequences, we observed the provision of information using patient-centered communication (GINFO+) and structure session which is the counselor either explaining the therapeutic agenda or attempting to transition to a new topic or session content (SS). Sometimes, counselor acknowledged the clients' communication or an off topic comment (SO). We also observed that the affirm and OQECML+ are consistent with MI theory as an effect for eliciting positive change talk or commitment language. For unsuccessful sequences, It can be seen that providing advice using non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-) which is heading in the wrong direction therapeutically. Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.

%It can be seen that most successful sequence starts with a summary of the discussion or structured statement. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation. Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.

>>>>

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. Sequence classification methods are utilized for classifying temporally ordered sequences. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}. In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease. Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.

@@

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information is learned from the dataset, and used as a ``model'' to classify sequences.

@@

Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

%This work is the next step in a series of experiments to develop the [informatics based methods described above]. 
In our previous work \cite{kotov2015interpretable, hasan2016study}, we explored several machine learning methods for automatic annotation of clinical interview fragments with a specialized codebook containing a large number of patient and provider behavior codes \cite{carcone2013provider}. In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers.
We propose Recurrent Neural Networks with two baseline probabilistic methods, MC and HMM, to identify patient-provider communication sequences that are likely to elicit the desired patient behavioral response (i.e. change talk or commitment language) and to dynamically estimate the likelihood of observing this desired response at any point during a clinical interview based on all coded previous patient-provider communication exchanges in the same interview. While there have been some previous qualitative studies of patient-provider dialog in a clinical setting \cite{eide2004physician}, there have been no previous work on computational modeling of annotated patient-provider communication (PPC) exchanges and predicting the desired patient behavior in a context of motivational interviews.

@@

Our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.85\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.15\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data level and algorithmic level. In this study, we utilized under and oversampling methods at data processing level for balancing the adolescent obesity data set. Synthetic Minority Over Sampling Technique (SMOTE) is the most widely used method for oversampling an imbalanced dataset, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. We generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, the undersampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.

A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive change talk at the end. It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively. Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors. For this reason, we only consider those models that can capture the past events for the classification of behavior sequences.\\

@@

\textbf {Hidden Markov Model (HMM)} is another probabilistic model used for modeling processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. Mathematically, HMM can be defined as the following: $\lambda = (A, B, \pi)$.
\begin{itemize}
\item A is an $N\times N$ state transition probability distribution matrix $A = \{a_{ij}\}$
\item B is an $N\times M$ matrix $B = \{b_j(k)\}$ with observation symbol probability distribution for each state 
\item $\pi$ is the initial state distribution vector $\pi = \{\pi_i\}$
\end{itemize}
Hence, N is a number of hidden states in the model and M is a number of distinct observations symbols per state, i.e. the discrete vocabulary size. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model deduces a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}.

@@

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. That RNN can capture long-term dependencies for predicting the future events is its main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit\cite{cho2014properties} or simply GRU. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication (TR)}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden layer of 15 LSTM or GRU nodes after embedding the given MYSCOPE codes from the input sequence. Then, we applied a softmax layer at each time-step for predicting the target which is the label of the entire sequence (successful or unsuccessful). This target replication (TR) strategy was used in several studies \cite{lipton2015learning, choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), total loss of our prediction models is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y^{(t)}$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Experimentally, it was determined that best performance achieved when we use  $\alpha=0.5$. Our model also contains several other hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with Adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

@@

\subsection*{\textit{Most likely communication sequences}}
Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. In successful sequences, we observed the provision of information using patient-centered communication (GINFO+) and structure session which is the counselor either explaining the therapeutic agenda or attempting to transition to a new topic or session content (SS). Sometimes, counselor acknowledged the clients' communication or an off topic comment (SO). We also observed that the affirm and OQECML+ are consistent with MI theory as an effect for eliciting positive change talk or commitment language. For unsuccessful sequences, It can be seen that providing advice using non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-) which is heading in the wrong direction therapeutically. Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.
%It can be seen that most successful sequence starts with a summary of the discussion or structured statement. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation. Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
